Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-08 paper/main.tex
start working on section 6

@@

@@

It is also possible to incorporate boolean conditions into a
bound to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre- and postconditions to take non-negative numbers or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

\end{lstlisting}
\end{minipage}

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \mbox{logical state invariant } \{ \code{na} = \#_1(a)} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1 && na > 0) {
    a[x]=0;
    na--; tick(1);
    x++; }
  if (x < N) {
    a[x]=1;
    na++; tick(1); }
  k--;
}

\end{lstlisting}
\end{minipage}

\caption{Example of assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.  The bound derived automatically is
  $2|[0,\code{k}]| + |[0,\code{na}]|$.
  }
\label{fig:xmplincaux}
\end{figure}
%
\begin{figure}
\begin{lstlisting}
${\color{blue}%
  \mbox{logical state invariant }
  \{ \code{lg} \ge \log_2(h-l) \cdot M_\code{bsearch} \} }$
bsearch(x,l,h,lg) {
  if (h-l > 1 && lg > 0) {
    m = h + (h-l)/2;
    lg--; if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h,lg);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We use the tick metric to model the stack consumption of the program.
  Before a function call to \code{bsearch}, a tick statement pays for
  the space used by stack frame of the function.  The space is recoverd
  after the call.
  The bound derived automatically is $|[0,\code{lg}]|$.
  }
\label{fig:xmplbsaux}
\end{figure}
%
\iffull{
}{}
We can then use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, we can derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.

>>>>

\lstset{moredelim=[is][\underbar]{_}{_},keepspaces=true}

@@

\sectskip
\section{Logical State and User Interaction}
\label{sec:anno}
\aftersectskip

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant } \{ \code{na} = \#_1(\code{a})} \}$
while (k > 0) {
  x=0;
  while (x < N && a[x] == 1 _&& na > 0_) {
    a[x]=0; _na--;_
    tick(1);
    x++; }
  if (x < N) {
    a[x]=1; _na++;_
    tick(1); }
  k--;
}
\end{lstlisting}
\end{minipage}
\caption{Example of assisted bound derivation using logical state.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.  The bound derived automatically is
  $2|[0,\code{k}]| + |[0,\code{na}]|$.
  }
\label{fig:xmplincaux}
\end{figure}

\begin{figure}
\begin{lstlisting}[numbers=left]
${\color{blue}%
  \mbox{logical state invariant }
  \{ \code{lg} \ge \log_2(h-l) \cdot M_\code{bsearch} \} }$
bsearch(x,l,h_,lg_) {
  if (h-l > 1 && lg > 0) {
    m = h + (h-l)/2;
    _lg--;_ if (a[m]>x) h=m; else l=m;
    tick($M_\code{bsearch}$);
    l = bsearch(x,l,h_,lg_);
    tick($-M_\code{bsearch}$);
  } else return l;
}
\end{lstlisting}
\caption{Example of assisted bound derivation using logical state.
  We use the tick metric to model the stack consumption of the program.
  Before a function call to \code{bsearch}, a tick statement pays for
  the space used by stack frame of the function.  The space is recoverd
  after the call.
  The bound derived automatically is $|[0,\code{lg}]|$.
  }
\label{fig:xmplbsaux}
\end{figure}

While complete automation is often desirable, it cannot possibly
be universal since the problem is undecidable.  In this section
we present two techniques to derive complex resource bounds by
leveraging our automation.

When the resource bound of a program depends on the contents of the heap,
or is non-linear (e.g. logarithmic, exponential), we propose to
introduce \emph{logical state}.
This technique is inspired by XXX (speed) and consists in introducing
extra variables to the program that do not change its behavior, but help
the automatic tool to figure out a bound.

The systematic technique consists in the following steps. First,
a program $P$ that fails to be analyzed automatically is enriched
by extra variables $\vec x$ called logical variables to form a
program $P_l(\vec x)$.  Second, an initial value $\vec X(\sigma)$
for the logical variables is shown to be such that:
$$
\forall n\, \sigma\, \sigma'.\,
  (\sigma, P_l(\vec X(\sigma))) \Downarrow_n \sigma'
    {\implies}
  \exists n', n' \le n \land
  (\sigma, P) \Downarrow_{n'} \sigma'.
$$
Since the annotated program and the original one are often
syntactically close, the proof of this result goes
by simple induction on the resource-aware evaluation judgement.
Third, using existing automation tools, a bound $B(\vec x)$ for
$P_l(\vec x)$ is derived.  Finally this bound, instantiated with
$\vec X$, gives the final resource bound for the program $P$.

This idea is illustrated by the program on \pref{fig:xmplincaux}.
Parts of the code underlined are annotations that were added
to the original program text.  The top-level loop increments
a binary counter \code{k} times.  A naive analysis of
the algorithm yields the quadratic bound $\code{k}\cdot\code{N}$.
However, the algorithm is in fact linear and its cost is bounded
by $2\code{k} + \#_1(\code{a})$ where $\#_1(\code{a})$ denotes
the number of one entries in the array \code{a}.  Since this
number depends on the heap contents, no tool available
for C is able to derive the linear bound.  However, it can be inferred
by our automated tool if a logical variable \code{na} is introduced.
This logical variable is a reification, in the program, of the
number $\#_1(\code{a})$.  For example, on line 5 of the example
we are setting \code{a[x]} to 0 and because of the condition we
know that this array entry was 1.  To reflect this change on
$\#_1(\code{a})$, the logical variable \code{na} is decremented.
Similarly, on line 9, an array entry which was 0 becomes 1, so
\code{na} is incremented.  To complete the step 2 of the systematic
procedure described above, we must show that the extra test
\code{na > 0} at line 4 does not influence the loop behavior.  It
is done by proving inductively that $\code{na} = \#_1(\code{a})$
and remarking that since $\code{a[x]} = 1$ is true, we must have
$\#_1(\code{a}) > 0$, thus \code{na > 0} in the test. %  So the
% test is in fact semantically equivalent to \code{true}.

Another simple example is given on \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are underlined in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $\code{lg} \ge
\log_2(h-l)\cdot M_\code{bsearch}$ as invariant.  This trick
allows a simpler proof because, when working with integer
arithmetic, it is not always the case that $\log_2(x/2) =
\log_2(x)-1$.

@@

It is also possible to incorporate boolean conditions into a
bound to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre- and postconditions to take non-negative numbers or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$
\end{lstlisting}
\end{minipage}

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
We can then use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, we can derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.
